Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-29 paper/paper.tex
James's changes

\begin{abstract}
    The current state of the art in subgraph isomorphism solving involves using degree as a
    value-ordering heuristic to direct backtracking search. Such a search makes a heavy commitment
    to the first branching choice, which is often incorrect. To mitigate this, we introduce and
    evaluate a new approach, which we call ``solution-biased search''. By combining a
    slightly-random value-ordering heuristic, rapid restarts, and nogood recording, we design an
    algorithm which instead uses degree to direct the proportion of search effort spent in different
    subproblems. This increases performance by two orders of magnitude on satisfiable instances,
    whilst not affecting performance on unsatisfiable instances. Such a such search can also be
    parallelised in a very simple but effective way: across both satisfiable and unsatisfiable
    instances, we get a further speedup of over thirty from thirty six cores, or ninety five from
    ten distributed-memory hosts. Finally, we show that solution-biased search is also suitable for
    optimisation problems, by using it to improve two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. The problem is broadly applicable, with applications in areas
including in bioinformatics \cite{DBLP:journals/bmcbi/BonniciGPSF13}, chemistry \cite{o:Regin95}, computer vision
\cite{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, law enforcement
\cite{DBLP:journals/cacm/CoffmanGM04}, model checking \cite{DBLP:journals/tcs/SevegnaniC15},
malware detection \cite{DBLP:conf/dimva/BruschiMM06}, compilers
\cite{DBLP:conf/cgo/MurrayF12,DBLP:conf/cp/BlindellLCS15}, pattern recognition
\cite{DBLP:journals/ijprai/FoggiaPV14}, program similarity
comparison \cite{o:DallaPredaV17}, the design of mechanical locks
\cite{DBLP:journals/siamrev/VomelLBF17}, and graph databases
\cite{DBLP:journals/jair/McCreeshPST18}.

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a \emph{pattern} graph $\mathcal{P}$ to a \emph{target} graph $\mathcal{T}$, such that
adjacent vertices in $\mathcal{P}$ are mapped to adjacent vertices in $\mathcal{T}$ (including that
vertices with loops in $\mathcal{P}$ may only be mapped to vertices with loops in $\mathcal{T}$).
The \emph{induced} problem additionally requires that non-adjacent vertices are mapped to
non-adjacent vertices. The \emph{degree} of a vertex is the number of other vertices to which it is
adjacent.

@@

We begin with a set of baseline performance measurements. In the top
two plots \cref{figure:others} we show the cumulative number of instances solved over time for the
non-induced and induced problems respectively. We compare the Glasgow Subgraph Solver using
depth-first search (DFS) and with the modifications described in the remainder of this paper
(solution-biased search, SBS), the PathLAD variation of the LAD algorithm
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).  The plots show that our starting point
comfortably beats PathLAD, VF2, VF3 and RI, except for very low choices of timeout.

@@

A \emph{discrepancy} is where search goes against the advice of a value-ordering heuristic.
Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies; in such cases, the heavy commitment to early branching choices made by conventional
backtracking search can be extremely costly.

@@

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to
one of, for example, three target vertices of degree ten, two vertices of degree nine, or five of
degree two? When driving conventional backtracking search, the degree heuristic would pick one of
the vertices of degree ten, and we would commit all of our search effort to the exponentially large
search tree underneath it, not considering any other choice until this tree has been fully explored
and eliminated.  We will show that this is not a wise choice, and that instead, we should
\emph{commit equal search effort} to each of the three subproblems found by mapping to vertices of
degree ten.  And similarly, should we be certain that vertex of degree ten is so much better than a
vertex of degree nine that we should commit no effort to degree nine vertices until all the degree
ten subproblems have been explored? Or might it be better to commit, say, twice as much effort to
each degree ten subproblems as to each degree nine subproblems, and only a very small amount of
effort to the degree two subproblems?  To test this hypothesis, we will now introduce a new
alternative to backtracking search, which we call \emph{solution-biased search}.  This search is
made up of three components: a new slightly-random value-ordering heuristic, rapid restarts, and
nogood recording. The aim is to perform a complete search, but spending proportionally more time in
parts of the search tree that are preferred by the value-ordering heuristic.

@@

To estimate solution density, we need an estimate of how big different subproblems are likely to be,
and of how many solutions each subproblem is likely to contain. Of course, obtaining exact (or even
approximate) values for these figures is at least as hard as solving the problem in its entirety.
For pairs of Erd\H{o}s-R\'enyi
random graphs with large solution counts (i.e.\ chosen from within the ``easy satisfiable'' region
\cite{DBLP:journals/jair/McCreeshPST18}), we observe a linear relationship between subproblem size
and number of solutions. Thus, for graphs from this distribution, we need only an estimator of
subproblem size. Measurements also suggest that, for pairs of Erd\H{o}s-R\'enyi graphs, subproblems
under a target vertex of degree $d$ tend to contain a small constant times more search nodes than
those under a target vertex of degree $d - 1$.  This empirical analysis suggests that an estimator
that is exponential in $d$ will give our method the desired behaviour, at least for
Erd\H{o}s-R\'enyi graphs. We expect it may be possible to derive better estimators for particular
input classes, although over the full range of problem instances, we have verified that estimators
of the form $e^d$ substantially outperform polynomial and factorial weightings.

>>>>

\begin{abstract}
    The current state of the art in subgraph isomorphism solving involves using degree as a
    value-ordering heuristic to direct backtracking search. Such a search makes a heavy commitment
    to the first branching choice, which is often incorrect. To mitigate this, we introduce and
    evaluate a new approach, which we call ``solution-biased search''. By combining a
    slightly-random value-ordering heuristic, rapid restarts, and nogood recording, we design an
    algorithm which instead uses degree to direct the proportion of search effort spent in different
    subproblems. This increases performance by two orders of magnitude on satisfiable instances,
    whilst not affecting performance on unsatisfiable instances. This algorithm can also be
    parallelised in a very simple but effective way: across both satisfiable and unsatisfiable
    instances, we get a further speedup of over thirty from thirty six cores, or ninety five from
    ten distributed-memory hosts. Finally, we show that solution-biased search is also suitable for
    optimisation problems, by using it to improve two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. The problem is broadly applicable, with applications in areas
including bioinformatics \cite{DBLP:journals/bmcbi/BonniciGPSF13}, chemistry \cite{o:Regin95}, computer vision
\cite{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, law enforcement
\cite{DBLP:journals/cacm/CoffmanGM04}, model checking \cite{DBLP:journals/tcs/SevegnaniC15},
malware detection \cite{DBLP:conf/dimva/BruschiMM06}, compilers
\cite{DBLP:conf/cgo/MurrayF12,DBLP:conf/cp/BlindellLCS15}, pattern recognition
\cite{DBLP:journals/ijprai/FoggiaPV14}, program similarity
comparison \cite{o:DallaPredaV17}, the design of mechanical locks
\cite{DBLP:journals/siamrev/VomelLBF17}, and graph databases
\cite{DBLP:journals/jair/McCreeshPST18}.

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a \emph{pattern} graph $\mathcal{P}$ to the vertices of a \emph{target} graph
$\mathcal{T}$, such that adjacent vertices in $\mathcal{P}$ are mapped to adjacent vertices in
$\mathcal{T}$ (including that vertices with loops in $\mathcal{P}$ may only be mapped to vertices
with loops in $\mathcal{T}$).  The \emph{induced} problem additionally requires that non-adjacent
vertices are mapped to non-adjacent vertices. The \emph{degree} of a vertex is the number of other
vertices to which it is adjacent.

@@

We begin with a set of baseline performance measurements. In the top
two plots of \cref{figure:others} we show the cumulative number of instances solved over time for the
non-induced and induced problems respectively. We compare the Glasgow Subgraph Solver using
depth-first search (DFS) and with the modifications described in the remainder of this paper
(solution-biased search, SBS), the PathLAD variation of the LAD algorithm
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).  The plots show that our starting point
comfortably beats PathLAD, VF2, VF3 and RI, except for very low choices of timeout.

@@

A \emph{discrepancy} is where search goes against the advice of a value-ordering heuristic.
Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies. In such cases, the heavy commitment to early branching choices made by conventional
backtracking search can be extremely costly.

@@

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to
one of, for example, three target vertices of degree ten, two vertices of degree nine, or five of
degree two? When driving conventional backtracking search, the degree heuristic would pick one of
the vertices of degree ten, and we would commit all of our search effort to the exponentially large
search tree underneath it, not considering any other choice until this tree has been fully explored
and eliminated.  We will show that this is not a wise choice, and that instead, we should
\emph{commit equal search effort} to each of the three subproblems found by mapping to vertices of
degree ten.  And similarly, should we be certain that a vertex of degree ten is so much better than a
vertex of degree nine that we should commit no effort to degree nine vertices until all the degree
ten subproblems have been explored? Or might it be better to commit, say, twice as much effort to
each degree ten subproblems as to each degree nine subproblems, and only a very small amount of
effort to the degree two subproblems?  To test this hypothesis, we will now introduce a new
alternative to backtracking search, which we call \emph{solution-biased search}.  This search is
made up of three components: a new slightly-random value-ordering heuristic, rapid restarts, and
nogood recording. The aim is to perform a complete search, but spending proportionally more time in
parts of the search tree that are preferred by the value-ordering heuristic.

@@

To estimate solution density, we need an estimate of how big different subproblems are likely to be,
and of how many solutions each subproblem is likely to contain. Of course, obtaining exact (or even
approximate) values for these figures is at least as hard as solving the problem in its entirety.
For pairs of Erd\H{o}s-R\'enyi
random graphs with large solution counts (i.e.\ chosen from within the ``easy satisfiable'' region
\cite{DBLP:journals/jair/McCreeshPST18}), we observe a linear relationship between subproblem size
and number of solutions. Thus, for graphs from this distribution, we need only an estimator of
subproblem size. Measurements also suggest that, for pairs of Erd\H{o}s-R\'enyi graphs, subproblems
under a target vertex of degree $d$ tend to contain a small constant times more search nodes than
those under a target vertex of degree $d - 1$.  This empirical analysis suggests that an estimator
that is exponential in $d$ will give our method the desired behaviour, at least for
Erd\H{o}s-R\'enyi graphs. We expect it may be possible to derive better estimators for particular
input classes, although over the full range of problem instances, we have verified that exponential
estimators substantially outperform polynomial and factorial weightings.
